---
title: "Calculating p-value functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating p-value functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(tidyverse)
library(flipr)
ngrid_in <- 20
ngrid_out <- 100
B <- 1000
n1 <- 30
n2 <- 30
set.seed(1301)
x1 <- rnorm(n1, mean = 0, sd = 1)
x2 <- rnorm(n2, mean = 3, sd = 1)
y1 <- rnorm(n1, mean = 0, sd = 1)
y2 <- rnorm(n2, mean = 0, sd = 2)
z1 <- rnorm(n1, mean = 0, sd = 1)
z2 <- rnorm(n2, mean = 3, sd = 2)

generate_grid <- function(center_value, min_value, max_value, n) {
  stopifnot(center_value > min_value && center_value < max_value)
  c(
    seq(min_value, center_value, len = n / 2 + 1)[1:(n / 2)], 
    center_value, 
    seq(center_value, max_value, len = n / 2 + 1)[-1]
  )
}
```

The concept of $p$-value functions pertains to assessing the $p$-value
of a set of null hypotheses and to plot this $p$-value surface on the
domain defined by the set of null hypotheses. The idea behind is that if
such a $p$-value function is available, you can deduce from it point
estimates or confidence interval estimates for parameters used to define
the nulls or extract a single $p$-value for a specific null of interest
[@martin2017; @fraser2019; @infanger2019]. In particular, there is
another **R** package dedicated to $p$-value functions called
[**pvaluefunctions**](https://cran.r-project.org/web/packages/pvaluefunctions/index.html).

## P-value function for the mean

```{r fig.width=6, out.width="100%"}
test_param <- stats::t.test(x2, x1, var.equal = TRUE)
delta_min <- test_param$conf.int[1] - 1
delta_max <- test_param$conf.int[2] + 1

delta_pe <- mean(x2) - mean(x1)
delta_pe
delta_in <- generate_grid(delta_pe, delta_min, delta_max, ngrid_in)

null_spec_mean <- function(y, parameters) {
  delta <- parameters[1]
  y - delta
}

df <- tibble(
  delta = delta_in,
  two_tail = two_sample_pf(
    parameters = delta, 
    null_specification = null_spec_mean, 
    x = x1, 
    y = x2, 
    statistic = stat_t, 
    B = B, 
    seed = 1234, 
    alternative = "two_tail"
  ),
  left_tail = two_sample_pf(
    parameters = delta, 
    null_specification = null_spec_mean, 
    x = x1, 
    y = x2, 
    statistic = stat_t, 
    B = B, 
    seed = 1234, 
    alternative = "left_tail"
  ),
  right_tail = two_sample_pf(
    parameters = delta, 
    null_specification = null_spec_mean, 
    x = x1, 
    y = x2, 
    statistic = stat_t, 
    B = B, 
    seed = 1234, 
    alternative = "right_tail"
  )
)

delta_out <- generate_grid(delta_pe, delta_min, delta_max, ngrid_out)

data_t <- tibble(
  delta = delta_out, 
  two_tail = approx(df$delta, df$two_tail, delta)$y, 
  left_tail = approx(df$delta, df$left_tail, delta)$y, 
  right_tail = approx(df$delta, df$right_tail, delta)$y, 
) %>%
  pivot_longer(-delta)

data_t %>%
  ggplot(aes(delta, value, color = name)) +
  geom_line() +
  labs(
    title = "P-value function for the mean", 
    subtitle = "t-statistic", 
    x = expression(delta), 
    y = "p-value", 
    color = "Type"
  ) +
  geom_hline(
    yintercept = 0.05,
    color = "black",
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = mean(x2) - mean(x1),
    color = "black"
  ) +
  geom_vline(
    xintercept = test_param$conf.int,
    color = "black",
    linetype = "dashed"
  ) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), limits = c(0, 1))
```

## P-value function for the variance

```{r fig.width=6, out.width="100%"}
test_param <- stats::var.test(y2, y1)
rho_min <- 1e-3# sqrt(test_param$conf.int[1]) * 0.8
rho_max <- sqrt(test_param$conf.int[2]) * 1.2

rho_pe <- sd(y2) / sd(y1)
rho_pe
rho_in <- generate_grid(rho_pe, rho_min, rho_max, ngrid_in)

null_spec_sd <- function(y, parameters) {
  rho <- parameters[1]
  y / rho
}

df <- tibble(
  rho = rho_in,
  two_tail = two_sample_pf(
    parameters = rho, 
    null_specification = null_spec_sd, 
    x = y1, 
    y = y2, 
    statistic = stat_f, 
    B = B, 
    seed = 1234, 
    alternative = "two_tail"
  ),
  left_tail = two_sample_pf(
    parameters = rho, 
    null_specification = null_spec_sd, 
    x = y1, 
    y = y2, 
    statistic = stat_f, 
    B = B, 
    seed = 1234, 
    alternative = "left_tail"
  ),
  right_tail = two_sample_pf(
    parameters = rho, 
    null_specification = null_spec_sd, 
    x = y1, 
    y = y2, 
    statistic = stat_f, 
    B = B, 
    seed = 1234, 
    alternative = "right_tail"
  )
)

rho_out <- generate_grid(rho_pe, rho_min, rho_max, ngrid_out)

data_f <- tibble(
  rho = rho_out, 
  two_tail = approx(df$rho, df$two_tail, rho)$y, 
  left_tail = approx(df$rho, df$left_tail, rho)$y, 
  right_tail = approx(df$rho, df$right_tail, rho)$y, 
) %>%
  pivot_longer(-rho)

data_f %>%
  ggplot(aes(rho, value, color = name)) +
  geom_line() +
  labs(
    title = "P-value function for the standard deviation", 
    subtitle = "F-statistic", 
    x = expression(rho), 
    y = "p-value", 
    color = "Type"
  ) +
  geom_hline(
    yintercept = 0.05,
    color = "black",
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = sqrt(test_param$statistic),
    color = "black"
  ) +
  geom_vline(
    xintercept = sqrt(test_param$conf.int),
    color = "black",
    linetype = "dashed"
  ) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), limits = c(0, 1))
```

## P-value function for both mean and variance

Assume that we have two r.v. $X$ and $Y$ that differ in distribution
only in their first two moments. Let $\mu_X$ and $\mu_Y$ be the means of
$X$ and $Y$ respectively and $\sigma_X$ and $\sigma_Y$ be the standard
deviations. We can therefore write

$$
Y = \delta + \rho X.
$$

In this case, we have

$$
\begin{cases}
\mu_Y = \delta + \rho \mu_X \\
\sigma_Y^2 = \rho^2 \sigma_X^2
\end{cases}
\Longleftrightarrow
\begin{cases}
\delta = \mu_Y - \frac{\sigma_Y}{\sigma_X} \mu_X \\
\rho = \frac{\sigma_Y}{\sigma_X}
\end{cases}
$$

In the following example, we have $\delta = 3$ and $\rho = 2$.

```{r fig.width=6, fig.height=5, out.width="100%"}
delta_pe <- mean(z2) - sd(z2) / sd(z1) * mean(z1)
delta_pe
rho_pe <- sd(z2) / sd(z1)
rho_pe
test_param_mean <- stats::t.test(z2, z1, var.equal = FALSE)
test_param_var <- stats::var.test(z2, z1)
delta_min <- test_param_mean$conf.int[1] - 1
delta_max <- test_param_mean$conf.int[2] + 1
rho_min <- 1e-3
rho_max <- sqrt(test_param_var$conf.int[2]) * 1.2

null_spec_both <- function(y, parameters) {
  delta <- parameters[1]
  rho <- parameters[2]
  (y - delta) / rho
}

process <- function(delta, rho, x, y, B, combine_with) {
  two_sample_pf(
    parameters = map2(delta, rho, c), 
    null_specification = null_spec_both, 
    x = x, 
    y = y, 
    statistic = c("stat_mean", "stat_f"), 
    B = B, 
    seed = 1234, 
    combine_with = combine_with, 
    alternative = "two_tail"
  )
}

process_cmp <- compiler::cmpfun(process)

delta_in <- generate_grid(delta_pe, delta_min, delta_max, ngrid_in)
rho_in <- generate_grid(rho_pe, rho_min, rho_max, ngrid_in)
delta_out <- generate_grid(delta_pe, delta_min, delta_max, ngrid_out)
rho_out <- generate_grid(rho_pe, rho_min, rho_max, ngrid_out)

df_fisher <- crossing(delta = delta_in, rho = rho_in) %>% 
  mutate(pvalue = process_cmp(
    delta = delta, 
    rho = rho, 
    x = z1, 
    y = z2, 
    B = B,  
    combine_with = "fisher"
  ))
df_fisher <- akima::interp(
  x = df_fisher$delta, 
  y = df_fisher$rho, 
  z = df_fisher$pvalue, 
  xo = delta_out, 
  yo = rho_out
)
pvalue_fisher <- t(df_fisher$z)

df_tippett <- crossing(delta = delta_in, rho = rho_in) %>% 
  mutate(pvalue = process_cmp(
    delta = delta, 
    rho = rho, 
    x = z1, 
    y = z2, 
    B = B,  
    combine_with = "tippett"
  ))
df_tippett <- akima::interp(
  x = df_tippett$delta, 
  y = df_tippett$rho, 
  z = df_tippett$pvalue, 
  xo = delta_out, 
  yo = rho_out
)
pvalue_tippett <- t(df_tippett$z)

fig <- plotly::plot_ly(
  x = delta_out, 
  y = rho_out, 
  z = pvalue_fisher
) %>%
  plotly::add_surface()
fig %>% 
  plotly::layout(scene = list(
    xaxis = list(title = "delta"), 
    yaxis = list(title = "rho"), 
    zaxis = list(title = "p-value")
  ))
```

```{r fig.width=6, fig.height=5, out.width="100%"}
fig2_fisher <- plotly::plot_ly(
  type = "contour", 
  x = delta_out, 
  y = rho_out, 
  z = pvalue_fisher
)
fig2_fisher %>% 
  plotly::layout(
    xaxis = list(title = plotly::TeX("\\delta")), 
    yaxis = list(title = plotly::TeX("\\rho"))
  ) %>% 
  plotly::config(mathjax = "cdn")
```

```{r fig.width=6, fig.height=5, out.width="100%"}
fig2_tippett <- plotly::plot_ly(
  type = "contour",
  x = delta_out,
  y = rho_out,
  z = pvalue_tippett
)
fig2_tippett %>%
  plotly::layout(
    xaxis = list(title = plotly::TeX("\\delta")),
    yaxis = list(title = plotly::TeX("\\rho"))
  ) %>%
  plotly::config(mathjax = "cdn")
```

```{r fig.width=6, fig.height=5, out.width="100%"}
fig3_fisher <- plotly::plot_ly(
  type = "contour", 
  x = delta_out, 
  y = rho_out, 
  z = pvalue_fisher, 
  autocontour = FALSE,
  contours = list(
    start = 0.05,
    end = 1,
    size = 0.05
  )
)
fig3_fisher %>% 
  plotly::layout(
    xaxis = list(title = plotly::TeX("\\delta")), 
    yaxis = list(title = plotly::TeX("\\rho"))
  ) %>% 
  plotly::config(mathjax = "cdn")
```

```{r fig.width=6, fig.height=5, out.width="100%"}
fig3_tippett <- plotly::plot_ly(
  type = "contour", 
  x = delta_out, 
  y = rho_out, 
  z = pvalue_tippett, 
  autocontour = FALSE,
  contours = list(
    start = 0.05,
    end = 1,
    size = 0.05
  )
)
fig3_tippett %>% 
  plotly::layout(
    xaxis = list(title = plotly::TeX("\\delta")), 
    yaxis = list(title = plotly::TeX("\\rho"))
  ) %>% 
  plotly::config(mathjax = "cdn")
```

## References
